Dispersion Distance and the Matter Distribution of the Universe in Dispersion Space

### 2015-06-04 writeup/dm_space_distortions.tex
rename file.

    The recently discovered fast radio bursts (FBRs) are a new class of radio
    transient that, due to their high dispersion from intervening plasma,
    have been interpreted as
    originating from cosmological distances.  The rate of fast radio bursts
    has been estimated to be several thousand over the whole sky per day.  If
    cosmological, the sources of these events should trace the large-scale
    structure of the
    Universe.
    We show here that, even without redshifts, these bursts can be used to
    study the three-dimensional cluster of matter in the Universe.
    This is because the dispersion of radio waves as they travel through the
    Universe, like redshift, can be
    used as a proxy for radial distance. 
    Due to inhomogeneities in
    the plasma's density, dispersion is an imperfect proxy for radial
    distance and we show that this leads to calculable dispersion-space
    distortions in the apparent clustering. We calculate the dispersion-space
    power spectra for a simple model of electron and FRB clustering and show that
    the clustering signal could be detected in a survey of $10^4$ FRB
    events. This rate is \ldots as is expected to be produced by the Canadian Hydrogen Intensity
    Mapping Experiment.

\end{abstract}

@@

Large-scale structure is usually studied through its two-point statistics, most
commonly the power spectrum, $P(k)$.
Unlike in redshift-space distortions, Equation~\ref{e:delta_s} does not have a
simple form in harmonic space. The equation's third term couples harmonic modes
and thus the two-point statistics cannot be phrased as a simple power spectrum.
We will instead use $C^{ss}_\ell(\chi,\chi')$, which is
the cross-correlation angular power spectrum of the DM-space over density,
on shells at $\chi$ and $\chi'$:
\begin{align}
    &\delta_{\ell\ell'}\delta_{mm'}C^{ss}_\ell(\chi, \chi') = \nonumber \\
    & \quad
        \left\langle
        \int\ud\Omega Y_{\ell m}(\hat n)
       \delta_{fs}(\hat n \chi, \chi)
        \int\ud\Omega' Y_{\ell' m'}(\hat n')
        \delta_{fs}(\hat n' \chi', \chi')
        \right\rangle.
\end{align}
The first two terms in Equation~\ref{e:delta_s} are stationary. For stationary,
isotropic, tracers $x$ and $y$ we have 
$\langle \delta_x(\vec k, \chi) \delta_y(\vec k', \chi) \rangle = (2\pi)^3
\delta^3(\vec k - \vec k') P_{xy}(k, \chi)$.  If for a moment we ignore time
dependence, the angular cross-power spectrum of such tracers is
\begin{align}
    %\delta_{\ell\ell'}\delta_{mm'}C^{xy}_\ell(\chi, \chi')
    %&= \int\ud\Omega\ud\Omega'Y_{\ell m}(\hat n) Y_{\ell' m'}(\hat n')
    %    \langle \delta_x(\hat n \chi) \delta_y(\hat n' \chi') \rangle,
    %    \\
    C^{xy}_\ell(\chi,\chi')
    &= \frac{2}{\pi}
\int_0^\infty\ud k k^2 j_\ell(k\chi) j_{\ell}(k\chi')P_{xy}(k).
\end{align}
In reality the power spectrum evolves on the order of a Hubble time.
The angular cross correlations will be very small unless $\chi$ and $\chi'$ are
within a few correlation lengths of one another, roughly a hundred
megaparsecs.  The evolution of the power-spectrum is negligible over these time
differences which leads to a straight-forward way to include the evolution of
the perturbations:
\begin{align}
C^{xy}_\ell(\chi,\chi') 
    \approx \frac{2}{\pi}
    \int_0^\infty\ud k k^2
    j_\ell(k\chi) j_{\ell}(k\chi')
    P_{xy}(k,(\chi + \chi')/2),\nonumber &\\
    |\chi - \chi'| \ll 1/aH.&
\end{align}

>>>>

    We propose that `standard pings', brief broadband radio impulses, can be
    used to study the three-dimensional clustering of matter in the Universe
    even in the absence of redshift information.  The dispersion of radio waves
    as they travel through the Universe, like redshift, can be used as a
    cosmological distance measure.  Due to inhomogeneities in the electron
    density along the line-of-siight dispersion is an imperfect proxy for
    radial distance and we show that this leads to calculable dispersion-space
    distortions in the apparent clustering of sources. Fast radio bursts (FRBs)
    are a new class of radio transients that are the prototypical standard
    ping, and due to their high observed dispersion from intervening plasma,
    have been interpreted as originating at cosmological distances.  The rate
    of fast radio bursts has been estimated to be several thousand over the
    whole sky per day, and if cosmological, the sources of these events should
    trace the large-scale structure of the Universe.  We calculate the
    dispersion-space power spectra for a simple model where electrons and FRBs
    are biased tracers of the large-scale structure of the Universe and show
    that the clustering signal may be detectable in a survey with as few as
    $10000$ events.  Upcoming wide-field radio telescopes, like the Canadian
    Hydrogen Intensity Mapping Experiment (CHIME), are expected to detect...
\end{abstract}

@@

Large-scale structure is usually studied through its two-point statistics, most
commonly the power spectrum, $P(k)$.
Unlike in redshift-space distortions, Equation~\ref{e:delta_s} does not have a
simple form in harmonic space. The equation's third term couples harmonic modes
and thus the two-point statistics cannot be phrased as a simple power spectrum.
We will instead use $C^{ss}_\ell(\chi,\chi')$, which is
the cross-correlation angular power spectrum of the DM-space over density,
on shells at $\chi$ and $\chi'$:
\begin{align}
    &C^{ss}_\ell(\chi, \chi') =
    \int\ud\Omega\ud\Omega' Y_{\ell m}(\hat n) Y^*_{\ell m}(\hat n')
    \langle \delta_{fs}(\hat n \chi) \delta_{fs}(\hat n' \chi')
        \rangle.
\end{align}
The first two terms in Equation~\ref{e:delta_s} are stationary. For stationary,
isotropic, tracers $x$ and $y$ we have 
$\langle \delta_x(\vec k, \chi) \delta_y(\vec k', \chi) \rangle = (2\pi)^3
\delta^3(\vec k - \vec k') P_{xy}(k, \chi)$.  If for a moment we ignore time
dependence, the angular cross-power spectrum of such tracers is
\begin{align}
    %\delta_{\ell\ell'}\delta_{mm'}C^{xy}_\ell(\chi, \chi')
    %&= \int\ud\Omega\ud\Omega'Y_{\ell m}(\hat n) Y_{\ell' m'}(\hat n')
    %    \langle \delta_x(\hat n \chi) \delta_y(\hat n' \chi') \rangle,
    %    \\
    C^{xy}_\ell(\chi,\chi')
    &= \frac{2}{\pi}
\int_0^\infty\ud k k^2 j_\ell(k\chi) j_{\ell}(k\chi')P_{xy}(k).
\end{align}
In reality the power spectrum evolves on the order of a Hubble time.
The angular cross correlations will be very small unless $\chi$ and $\chi'$ are
within a few correlation lengths of one another, roughly a hundred
megaparsecs.  The evolution of the power-spectrum is negligible over these time
differences which leads to a straight-forward way to include the evolution of
the perturbations:
\begin{align}
C^{xy}_\ell(\chi,\chi') 
    \approx \frac{2}{\pi}
    \int_0^\infty\ud k k^2
    j_\ell(k\chi) j_{\ell}(k\chi')
    P_{xy}(k,(\chi + \chi')/2),\nonumber &\\
    |\chi - \chi'| \ll 1/aH.&
\end{align}
